\chapter{Conclusion}


It is generally accepted that SLEs are becoming more frequent globally thus far resilience research often focuses on, highly vulnerable locations. However looking more closely at an area which is regarded as less can help create a reference point or target for a locations place-based resilience.
\paragraph{}

Much of Trondheim's infrastructure and population is situated within the coastal zone (figure \ref{fig:research_site}), as is common in Norway. To understand Trondheim's changing resilience to sea level extremes, its technological, natural and social systems resilience were considered.  Projected resilience was viewed as a projected outcome, which can be determined by analysing the systems which impact the risk (section \ref{theory-resilience}). In this case social, natural and technological, systems impacting the risk of sea level extremes. 
\paragraph{}

Literature reviews, investigation of the coastline, analysis of city plans and planning permissions were used to determine Trondheim's natural and technological systems resilience. Online surveys of stakeholders were used to aid the determination of Trondheim's social system resilience, specifically investigating the key variable of awareness (section \ref{data-collection}). Awareness is an important theme in this thesis falling under the broader concept of local knowledge, which in turn falls under the concept of social system resilience (section \ref{theory-resilience}). This framework was designed based on feedback received from a focus group and a pilot study that carefully considered the framing of textual and visual communication, particularly the use of edited photographs to display likely future scenarios of sea level extremes in Trondheim (section \ref{data-collection}). Four places within Trondheim - Skansen, Grillstad, Brattøra and Nidelva - were selected as study areas to investigate the city's resilience, due to their physical vulnerability to SLEs and large daily throughput of people which included a broad representation of society.

\paragraph{}
This thesis determines Trondheim's projected resilience to SLEs in the 2022-2050 period as high (section \ref{RQ1-findings}), while the projected resilience for 2050-2100 is determined to be very high (section \ref{RQ2 - findings}). Resilience is highly dynamic and over the extended time frame considered in this thesis (2022-2100) these results are likely to change. Hence, the requirement for designing a framework for, repeated measurements to determine the trend of a places resilience as was attempted here. 

\paragraph{}
Stakeholders were deemed somewhat aware about changes to SLEs (section \ref{RQ3 - finding}). The factors deemed to have impact on stakeholders' awareness of SLEs are whether they are residents, or whether they utilise the sources of family, newspapers and formal education for their information on climate change. This result enforces the importance of framing the textual and visual communication for the designated audience as outlined in the communication design section of the methods (section \ref{com-design}) as this can impact awareness and perception.  
\paragraph{}


Whether the surveys utilising visual simulations of sea level extremes to determine Trondheim's social system resilience to SLEs can also be used to improve awareness, hence improving resilience, would be an interesting avenue for future research. Also worth looking at is the impact on the research of including sectors of society who are often excluded from such research, for example by only utilising volunteer based stakeholder workshops. These sectors can be viewed as possibly the most vulnerable and include groups such as migrants, short term residents, tourists, youth and the immune compromised. Improving resilience is a requirement of the Sendai framework and the United Nations Sustainable Development Goals 11 and 13. This thesis has shown that carefully designed surveys may be a useful tool in a framework designed to repeatedly measure resilience to see whether these goals are being met. By considering the trend, we can determine whether Trondheim is in line with the United Nations Sustainable Development Goals (UN SDG’s) 11 and 13 - requirement of improving resilience of human settlements.


%quotes from thesis to consider highlighting in conclusion

%Stakeholder surveys conducted in Trondheim coastal neighbourhoods indicate better awareness of the risk of sea level extremes for the period 2050 to 2100 than 2022 to 2050. An  investigation of the link between local knowledge and academic models showed that residency was the most important factor affecting participants' awareness of sea-level extremes.  By combining social resilience with natural and technological resilience it allows for a more a holistic understanding of changing resilience to SLEs within Trondheim.

%This thesis was developed to assist in the creation of such a framework, which could quickly and inexpensively determine a place's resilience to changing SLEs. Determining an overall understanding of a place's resilience requires connecting local knowledge with academic knowledge.

%If the answers do not match the models this highlights a need for future research to find out why this is not the case.

%Local knowledge is often highlighted as a key facet of community resilience (\cite{setten_we_2019}). The assumption is that higher levels of local knowledge, particularly higher levels of awareness about natural systems and natural hazards, increases community resilience.

%Awareness of the changing tides and risk of storm surges is considered here as an aspect of local knowledge and is one of the key factors investigated to determine the resilience of Trondheim to SLEs. 

% The Safety class 1 of the building regulations TEK10/TEK17 as used since 2016, corresponds with the 20 year storm surge (\cite{tides_high_2022}) and the increasing number of building which compile with these regulations are part of the reason for higher projected resilience in 2050-2100 than 2022-2050.

% This newer conceptualisation of resilience allows for greater focus on the integration of local knowledge, but this facet is not always included (\cite{moser_turbulent_2019}).

%\cite{cutter_place-based_2008} introduced the Determination of Resilience of Place
%Resilience is here considered as the ability to return to normality as quickly as possible after an event (\cite{cutter_place-based_2008}).

% By 2090, 60 buildings in Trondheim are projected to be permanently flooded.

%Surprisingly, subjects predicted the future better than the past when asked about sea level changes in Trondheim (figure \ref{fig:slr_past} and figure \ref{fig:slr_future}). These patterns of awareness level may be explained by consulting narratives of SLEs in the media, where there is a focus on rising sea levels. Hence, the subjects expected that sea levels have already risen in Trondheim (figure \ref{fig:slr_past}), contrary to the recorded evidence from \cite{tides_high_2022}.


%How memory is formed and how it impacts awareness and resilience is highly debated. There is evidence that institutions, particularly the media, have the potential to influence memory formation and recollection (\cite{de_guttry_expiry_2022})

%This highlights an idea that if the SLE does not impact personal property or their daily activity then it will not affect them at all. 

%Future research could include questions to gain a greater understanding of how stakeholders view the role of the welfare state or insurance policies in their perception of this risk. 