

\chapter{Discussion of Results}
The results of the previous chapter are evaluated here and then these findings are put in the context of the theory. After this the findings are used to answer the three research questions. 



\section{Technological Systems Resilience}
The three systems of resilience based of \cite{cutter_place-based_2008} are highly interactive. Distinguishing technological systems, from natural systems for the risk associated with SLEs is particularly difficult in Trondheim due to the long history of the coastline being shaped by its population. Table 2.1 in the background  displays how each of the research sites coastline has been artificially altered. The land use and land type dictate the physical vulnerability (\cite{opach_seeking_2020}).
\paragraph{}

With current city organisation, only once SLEs reach the height of 269cm, which is the projection for the 20 year return height in 2090, is a critical building due to be impacted (\cite{kartverket_se_2020}). This does not mean that there is no impact to infrastructure, but that it is likely that the technological systems in Trondheim are able to quickly return to normality for SLEs below this level. However Trondheim Municipality's plans indicate that the research sites should be prepared for an SLE 4.87m by 2100 (\cite{hanssen_saksframlegg_2013}).
\paragraph{}
Once 4m SLE is reached 1,162 buildings will be flooded, 5 of these being critical buildings (table 5.1). If the current organisation of Trondheim remains this is projected to occur in the period 2080 to 2100. 2080 is 58 years, meaning there is a good chance of the city layout shifting and even if it remains the same there is not a a high likelihood of this level of flooding impacting the city.  
\paragraph{}
The ability to return to normality is only possible for sites which are temporally flooded. By 2090 60 buildings in Trondheim are projected to be permanently flooded. These buildings are not considered as possessing resilience, but they are a small part of the overall technological system. The impact of permanent flooding and nuisance flooding and how it interacts with Norway's system of insurance is discussed in the section on risk perception. While these floods may have minimal effects on the technological system resilience they are likely to affect the economics of the sites, including due to impacts to insurance premiums (\cite{cutter_community_2020}). 
\paragraph{}
Residency patterns are mainly considered under social systems of resilience. However the size, spread and density of building impacts resilience to natural hazards. In the case of resilience to SLEs this is particularly the case if as has been suggested there is more intrastate and residency moved onto reclaimed land such as seen in Grillstad. If this occur ed this could decrease Trondheim's technological system resilience.
\paragraph{}
Other infrastructure in Trondheim is projected to be resilient to future SLEs, but the drainage network and tunnels are at greater risk (\cite{hanssen_saksframlegg_2013}). This is due to there lower, often subterranean position. The drainage network is perhaps the greatest weakness in the technological system resilience, but there is ongoing work to improve it (\cite{hanssen_saksframlegg_2013}). This ongoing improvement again highlights the need for repeated measurement to understand a places trend in resilience. While vulnerability is here considered as a value, resilience is an ongoing, dynamic process.
\paragraph{}
The interpretation of whether a system is resilient is dependant on what the acceptable level of risk is. Is the level of resilience acceptable to stakeholders is a different question than whether the system is resilient. This could be an interesting course for future research and likely of particular interest to policy makers.
\paragraph{}
Overall the technological resilience of the city for the period 2022 to 2050 is considered very high. The technological system resilience of the city for the period 2050 to 2100 is considered high. This is because even with the increased risk, the building regulations implemented in 2016 are preparing for this risk. They are not designed so that no new infrastructure will be flooded, but that it will not be too negatively impacted by the flooding. In other words it should be able to quickly return to normality. 


\section{Natural Systems Resilience}
The projections of SLEs in this project were based of the fifth assessment of IPCC RCP 8.5 emission scenario: this scenario is considered the likely emission projection for business as normal (\cite{hanssen-bauer_climate_2017}). This scenario was chosen due to the uncertainty of calculation on SLEs, particularly those associated with storm surges. By potentially overestimating the impact of SLEs it is hope that this prevents overestimating the resilience of places to SLEs. Furthermore there is a potential that emission pattern will match those of RCP 8.5. 
\paragraph{}

For Trondheim it is projected that a sea level change of -1 to +21 cm will occur between 2022 and 2050, then a further sea level rise of 22 to 77 cm between 2050 and 2100 (\cite{hanssen_saksframlegg_2013}). The most important detractor for Trondheim's natural systems resilience is due to storm surges. These are expected to be particularly problematic due to accumulation effect of being place in a long, deep fjord (Figure 1.1). While Tronndheimsfjord currently gives Trondheim city protection against storms washing in from the North sea, it may exaggerate the effects of storm surges (\cite{hanssen_saksframlegg_2013}). 
\paragraph{}
The models of SLEs have significantly improved over that last decades. Switching from a bath tub model to including the aspects of isostatic uplift which is particularly important for locations in Norway. The results from the survey indicate that the subjects have a lack of awareness of the impacts of isostatic uplift in Trondheim. Trondheim's sea level has actually decreased over the last 30 years. There is a slight uncertainty is this value (15cm +-5cm), but there is confidence that the sea level has dropped. Why only three subjects correctly identified this trend (Figure 5.22) may highlight how people view there location as staying the same.

\paragraph{}
It was surprising that over ten times as many subjects answered in a way that corresponds with models for (\cite{kartverket_se_2020}) for the question of how will sea level change in the next 30 years than for how have sea levels changes in the last 30 years (Figure 5.21 and Figure 5.22). However this may be the blinders of being involved in this research. When consulting narratives of SLEs present in the media there is a focus on rising sea levels. Perhaps the subjects recognised that the world is getting "wilder, wetter and weirder" (\cite{rod_integrated_2012}) and they push this change backwards into there memory of how sea levels in Trondheim have changed. 
\paragraph{}
An alternative view would be to question the recording of this change in SLEs. There is a long history of undervaluing local knowledge in the physical science. However sea level is particularly hard to observe in Trondheim due to the continuous impact of tides and weather. It is thought here that this is a stronger argument for the differences between the models and the subjects responses. Determining sea level change requires repeated measurements and the calculation of the impact of tide and weather.This process needs to occur for many years and the measured change is rather small and would be difficult for anyone to spot. There was an attempt made at making sure the information presented in the survey on sea level change was as accurate as possible. The tide gauge in Trondheim was moved in 1990 and it was for this reason 30 years was selected as the time frame for these numeric questions on sea level change, to minimise measurement errors. 

\paragraph{}
Subjects had much higher "correct" answer rates for the questions which utilised edited photographs of simulated SLEs, rather than purely numeric. This difference in the question format likely played a large role as did that the purely numeric questions had 7 choices, while the edited photographs only had 4. When it comes to daily observance of SLEs a photograph is much closer to the lived experience than a numeric value. Numeric values were included in these questions to allow for that knowledge still to be displayed and allow those with poor eyesight to still distinguish between the photographs, even if using an e-reader.
\paragraph{}
Natural systems resilience to SLEs in Trondheim is considered here as high for both 2022 to 2050 and 2050 to 2100. The flooding projected from SLEs will not prevent a quick return to normality. However, the changing climate and ecological systems when taken as whole are much less resilient. The appropriateness of using single-hazard analysis of resilience is outlined in the section on discussion of framework.



\section{Social Systems Resilience}
The focus of social systems resilience thus far has been on local knowledge, awareness and community resilience. There are other facets of social system resilience including economic and cultural resilience. Cultural resilience is further discussed in the section on risk perception. Economic systems are fragile to disasters (\cite{head_comment_2020}). Currently Norway and Trondheim has strong economic systems, with large amounts of buffer. However repeated disasters can decrease this resilience. Of particular interest for SLEs is how insurance and resilience interact.
\paragraph{}
As mentioned earlier there is evidence that many stakeholders (especially residents) do not consider having insurance as being connected to resilience. How insurance impacts the individual who has experienced the direct impact of damaged or lost property is discussed in \cite{whitmarsh_are_2008}. It has also been highlighted that economic resilience can be impacted as lower levels of flooding impact than other systems (\cite{cutter_community_2020}). Currently in Norway household flooding insurance is not dependent on location, but it is predicted here that this would not continue if there is places which are getting repeatedly flooded.  How insurance will deal with increasing risks from natural hazards caused by the changing climate is of worldwide concern.

\paragraph{}
The other aspects of social system resilience are explored in the subsections below.


\subsection{Place and Language}
Table 5.2 shows that the most popular place for responses was Nidelva. Nidelva is the most central of the locations and has the greatest daily throughput of people. It also includes perhaps the most iconic views of Trondheim. Due to the prioritisation of video on Facebook and Instagram a short video displaying the changing SLEs projected for Nidelva was attached to each social media post. This may be part of the reason that Nidelva was the most popular place for subjects to respond on.  Next popular was Grillstad, perhaps due to the recognition by residents that the area could be severely influenced by flooding from SLEs. The researcher has few personal connections to this place, so this high level of responses was particularly interesting. Skansen and Brattøra are the next most responded to, both of these locations have significant commercial ventures. Brattøra in particular is dominated not by residency but by offices and industry. Perhaps the conduction of this survey in Summer decreased the number of responses due to the lack of office workers. Almost evenly split for each location was whether the survey was completed in Norwegian or English. English surveys had 66 response, while the Norwegian survey had 87 responses.
\paragraph{}
A key consideration in comparing the difference between responses in English and Norwegian is how much comparison is appropriate. Both surveys were written by the researcher and there was an attempt to keep the meaning of each questions as similar as possible. However there is no such thing as a perfect direct translation. The nuance of the terms is different and even when the words have the same meaning there is cultural differences in the conceptualisation of these terms. Also there is likely cultural differences between the subjects in their conceptualisation of risk and resilience. Furthermore the increased likely-hood of subjects responding in English having lived in another location first will influence there view of Trondheim's resilience. How language used impacts and individuals' view of risk and resilience is being worked on by ***.

\paragraph{}
Note that due to the set up of the survey it is unknown the migration status of the subjects. It is unknown if Norweigain or English is their first language. Due to the high level of people in Trondheim who can communicate in one of these languages having the survey only available in these languages was deemed appropriate. If this project was to be repeated in a different location, which languages should be used should be carefully considered.


\subsection{Interest Level in SLEs}
Awareness as a facet of local knowledge was the key variable within this project. The purpose of the survey was to allow for the determination of awareness in Trondheim of SLEs. An interesting comparison is the level of interest against awareness. There is evidence that community interest in the natural hazards, improves awareness and resilience (\cite{cutter_community_2020}. Interest level in SLEs does not appear to be a factor influencing Stakeholders awareness about SLEs in Trondheim. There is no linearity or other pattern observed in Figure 5.11. This raises questions about the assumption of resilience due to the presence of professionals within a community. However, there is a serious limitation: no marine workers were identified in the survey subjects. Subjects who ranked their interest in sea level as high (4) had the largest variation in determined awareness according to the number of answers they gave which matched models from \cite{kartverket_se_2020}.
\paragraph{}
Furthermore Interest level had no correlation with the physical vulnerability of the place the subjects chose to respond on. It was hypothesised that interest level would be dependent on this factor. The lack of professional interest (12) could be a limiting factor here to the findings on interest level, however there was 40 subjects with a high level of interest. The tendency for individuals to select the middle answer is  likely part of the reason why almost half the subjects selected that they have a medium interest in SLEs. One proposal is that those with higher or above levels of interest in SLEs, simply did not have knowledge about SLEs of the specific places in which the surveys were on. Trondheim has not had recent disasters associated with SLEs. How well high levels of academic knowledge relate to local knowledge is an ongoing aspect of study for resilience to natural hazards (\cite{lujala_role_2020}).
\paragraph{}



   

\subsection{Memory of SLEs and Length of Place-based Knowledge}
As shown in Figure 5.12 memory of SLEs in Trondheim is a very skewed factor. This is not surprising. Memory connects to length of knowledge and the length of knowledge for the majority of subjects does not extend far enough (see Figure 5.13) to include each of these SLEs events in Trondheim. It was highlighted in the results that 96 subjects had a length of knowledge of greater than five years. This time period includes the SLE event occurring in February 2020, which only 51 subjects responded that they remembered. There are many potential reasons whyy 45 subjects who had a long enough length of knowledge did not remember this event. 
\paragraph{}
There is the first potential that they were unaware of the event occurring, perhaps due to lack of personal impact, the short lived nature of the event, or not being present in Trondheim at the time. Secondly peoples understanding of timelines can be limited. Even if they remember the event the date may not be well associated. This is part of the reason why specific dates were not mentioned in this question, just years and months. The media precursor to this event, stated that Trondheim should not worry about flooding (\cite{baisotti_danger_2020}). The SLE for February 2020 in Trondheim was 215cm, exceptionally high, but very little was reported upon it. How information sources, particularly newspapers influence stakeholders awareness is discussed in the next section on Information Access.
\paragraph{}
How memory is formed and how it impacts awareness and resilience is a highly debated field. There is evidence that institutions, particularly the media have the potential to have strong influences on memory formation and recollection (\cite{de_guttry_expiry_2022}). There is evidence that creating institutional memory of disasters due to natural hazards can help minimise the impact of the next one (\cite{de_guttry_expiry_2022}). However non of the SLEs that have occurred in Trondheim since 1950 are considered here as a disaster, but simply an event. Perhaps this influences the lack of memory of SLEs in Trondheim. 
\paragraph{}

The emotional relationship between people and place likely influences their memory of SLEs. It will also influence their views on risk to the place and the acceptable level of resilience for the place. A high emotional connection to a place may influence the fact that stakeholders often do not consider insurance as an acceptable form of resilience (Gonzalez-Riancho et al (2017) as cited in \cite{gerkensmeier_governing_2018}). Current insurance patterns in Norway mean that the insurance to flooding is the same no matter on how physically vulnerable to this risk the building is. This may change in the future and affect how people view this aspect.
\paragraph{}
Future research could consider the impact of place attachment, as discussed by \cite{ariccio_place_2021}, on resilience and acceptable levels of resilience of place according to stakeholders. There appears to be potential of utilising memory anchoring to increase social resilience to storm surges (\cite{de_guttry_expiry_2022}).  The theories of place attachment and memory anchoring highlight potential ways for a place to improve its social system resilience. 


\subsection{Information Access about Climate and Place}
Neither awareness all nor awareness photo appeared dependent upon any of the information sources about place  (Table 5.8). Awareness all appeared dependent on family. The researcher is aware that they have discussed this project with many members of their family and that this could have influenced the results. For this reason extra Kruskal Wallis tests, upon whether either forms of awareness were dependent upon survey access, were conducted. The results allowed for the acceptance of the null hypothesis that survey access (including due to connection to researcher) is independent of either determinations of awareness. Then the independence of survey access and whether the information source of family was also checked. Again these factors were deemed independent. These tests indicate that the fact that Awareness all, appears dependent on family can not easily be simplified into the narrative that this is only a factor for the researchers family. It does not totally clear the potential of this but opens up other potentials. For example it is known that people accept information about the changing climate better from sources they trust (\cite{corner_a_principles_2018}). 
\paragraph{}
After the preliminary trials of the main survey the first group who got access to the survey upon opening to the public were test engineers. This group was targeted to allow errors to be quickly noticed and fixed before the final results were impacted. The first 15 participants had access to 8 surveys, one of which included a typo and was missing education. The option of choosing "utddanning" (formal education) as where you learned about climate change was missing from two of the surveys. These errors were quickly remedied and a maximum of five percent of subjects could have been impacted. 
\paragraph{}
Awareness Photo appears dependent on the information sources of formal education and newspaper for information about the changing climate. The small limitation on this factor is that for the very first few hours of data collection one of the surveys was missing the option to select formal education. This was determined to impact a maximum of three responses and was quickly fixed. For this reason this error in data collection method is deemed to have no major impact on the results. 
\paragraph{}
Of more significant impact is that the first day of data collection did not have peer-reviewed publications as an option for information source about changing climate. This was only added after it was written in by respondents as an answer they would like to have. Which was further highlighted as a subject emailed the researcher about this via the email option presented on the website used to gather stakeholders and direct them to the Nettskjema surveys.  The first 15 respondent did not have peer-reviewed publications as an option in the question about information sources for changing climate. However two of these subjects had it added in later, because they wrote about this when given the opportunity for extra information to go to the researcher. To investigate the potential impacts of this all the subjects who completed the survey before this was an option had this selection artificially added to see if it impacted the Kruskal Wallis Test results. The result for this attempt to minimise impact of this miss of information collection is that the null hypothesis was still accepted. Due to theses extra results, that the information source of peer-reviewed publications about climate change is independent of awareness is still the conclusion made here, but there is a greater uncertainty to this result.
\paragraph{}


\subsection{Community Membership}
An important limitation in the study is that no subjects responded that they were marine-workers as displayed in Figure 5.14. Future research should target this group earlier in the data collection process. The lack of non-marine workers prevented analysis into whether their professional knowledge of the sea would translate into high determined awareness of SLEs. The use of emails was a major technique used to reach marine workers, the limiations of this technique are discussed in the next section on access to survey. 

\paragraph{}
The most popular community membership selected was resident with 70 out of 153 subjects choosing this option. Even with its high popularity this result was  surprisingly low as it was predicted that mainly residents would be interested in taking part in this research. Perhaps it was this concern that there would be a lack of non-residents that made the researcher overly focus on accessing non-residents and encouraging them to take part in the research. The high levels of lack of residents may also be due to the targeting of students. By many definitions of resident, lots of the students who participated in this project would also be considered residents. However the majority of subjects only selected one community membership. An important factor in interpreting Figure 5.14 is that there were only 221 responses for a subject group of 153. The mode number of responses to this question was one. That the majority of subjects only selected one response is also repeated in other questions where the subject could select multiple answers. 
\paragraph{}
The lack of students who also selected resident may be due to the tendancy to only select one answer. However, it may have been due to different conceptualisation of what it means to be a resident. Were subjects considering themselves a resident only if they resided in the specific place they responded on. Or did they consider themselves a resident if they lived in Trondheim. There is a potential that the conceptualisation of resident could have been linked to those who held Norwegian permanent residency. For future research perhaps an alternative term could be used, or who is considered a resident defined within the survey. Whether the preamble to these types of questions needed to mention please tick all that apply is also something that should be considered if this technique is to be repeated. 
\paragraph{}

\subsection{Access to Survey}
While access to survey does not appear to have influence on awareness of local knowledge, the results are important for the discussion on how to create a framework for determining resilience of a place. However almost all subjects only chose one access type. Just as with Figure 5.14 on community membership, Figure 5.15 on Access to Survey shows that the majority of respondents only chose a single answer. Once again this is a limitation on the conclusions that can be drawn from this result. Whether every subject who had a personal connection to the researcher chose that access method can be debated. Again terminology here is important, what does it mean to have a personal connection. Does a personal connection mean solely a family member or close friend or can it include acquaintances, colleagues or those with a shared group membership.  For the determination of awareness which considered all five possible questions, Family is a significant factor. This means that this determination of awareness appears dependent on family as a sources of observation. This result is important in the context that 25 subjects have a personal connection to the researcher.
\paragraph{}
Beyond the researchers personal connection to their family another personal connection is to the members of Trondheim Kajakk Klubb. Trondheim Kajakk Klubb were a targeted group due to their position as frequent water leisure users. The researcher is a member of this club which made it easier to access the members. However, it is believed that the members accessed the survey via social media, rather than access due to organisation of membership or due to personal connection to researcher. It is known that at lest seven subjects who could have selected all three of those responses, to the question on access to survey, only chose one. Again this is a limitation on the interpretation of the results. Though there may have been some impact, the results indicates that if this project was done in a location where the researcher does not have a connection would still yield results.
\paragraph{}
The least popular responses to the question on access to survey were place of employment, organisation of membership and email. Many firms have effective email filters in place which could have placed the email attempting to reach stakeholders into the spam. Furthermore the time period of collection was the summer of 2022, which is when many people take their holiday. This means that this was a poor time to access potential stakeholders via their email, place of employment or organisation membership. It also means that there were more visitors and tourists available as potential subjects. It is expected that if this survey was repeated in winter then this result would look very different. For this reason their is no conclusion given on the appropriateness of accessing stakeholders via place of employment, organisation membership or email.
\paragraph{}
That 70 subjects accessed the survey via the poster, indicates that this method was successful in reaching stakeholders. The design of the poster likely had significant impact on that. To see the poster used consult the appendix. To see the communication guidelines used consult the methods section. It is likely that posters were more successful at attracting stakeholders when placed in certain locations. There was a consideration of making sure that each qr-code used to access the posters was linked to the precise location of the posters. This was decided against as to print 50 different posters, which had the plastic coating needed to withstand the weather conditions in Trondheim for at least one month, was outwith the researchers budget. This investigation method could be interesting for future research. The next most frequent choice on how subjects accessed the survey was social media. When considering a framework to allow repeated measurements of resilience these two techniques used to access stakeholders is recommended here. 
\paragraph{}

\section{Perceived Risk}
This survey will have influenced the subjects perception of risk, it is unlikely that 97 subjects would have selected storm surges as a risk to infrastructure before carrying out this survey. The impacts of citizen science and perception of risk is expanded upon in the discussion of framework, including the high levels of perception that shoreline instability is a risk to infrastructure in the research sites.

\subsection{Risk Perception - Personal Impacts of SLEs and associated Flooding}
There was opportunity for subjects to give extra information on certain topics as can be seen in the appendix. From these responses narratives about perception of risk can be explored. For example their was a very varied response to "How would flooding associated with SLEs in this area affect you? If you would like to give more details on this, please write below." This has been grouped into narratives depending on the impact level the subjects gave to the previous question.
\paragraph{}

Subjects who felt they would have no impact who gave more details either stated that they lived far from the sea, or that they would be moving away soon. This highlights an idea that if the SLE does not impact items they own or their daily activity then it will not affect them at all. Nuisance flooding was not considered an issue, nor how flooding could impact key infrastructure for example transport connections. One subject gave details of how flooding did impact where they normally park their car, but as it wasn't parked during this event then they weren't impacted. 
\paragraph{}

Subjects who responded that flooding in this area would have mild impact and gave more information gave a wider range of responses than those who felt it would not impact them. The overall narrative was that flooding won’t affect where they live, but may impact their family and their activities. One curious response was from an individual who said that because they lived on a higher floor they would not be impacted. Living on a higher floor may provide some protection but in a extremely high sea level could still have impacts on daily activities. This continues the narrative that SLEs may impact my activities but as long as it doesn't negatively impact my possessions the impact can be considered mild. This raises questions about what it means to experience a flooding hazard. What it means to have direct experience can have varied definitions. 
\cite{whitmarsh_are_2008} defines direct flooding experience as those who have experienced forms of flood damage to their home, garden or vehicle within the last half decade. This is a particularly tight definition of direct experience and focuses on ownership. People who do not own a home, garden or vehicle can still be impacted by flooding and their experience may be very different.  
\paragraph{}
It is commonly assumed that direct experience makes people more concerned about climate change and its potential personal impacts, but this does not always appear to be the case (\cite{lujala_role_2020}). Perhaps some of this variation is due to variation in what is defined as direct experience. Further research could look into differing types of direct experience. 
\paragraph{}
Only once subjects responded with medium impact did a negative emotional response get expressed. Subjects discussed how flooding will affect places where they stored things, particularly at the harbour. Subjects who ranked flooding in this area as Significant impact gave similar response. The overall narrative expressed from the extra information given here is that subjects appear to believe that if SLEs do not affect the metres of where they live it will not affect them. 
\paragraph{}
None of the subjects who gave extra information mentioned insurance as a factor in their perception of personal impact level of flooding due to SLEs. Gonzalez-Riancho et al (2017 as cited in \cite{gerkensmeier_governing_2018}) says that stakeholders for the Wadden Sea Region a similarly developed area in Northern Europe, who has a longer history of negative flooding impacts from SLEs, reject that insurance can be considered as a form of disaster risk management. Perhaps the self-selecting stakeholders for Trondheim have similar views. How economists view the role of insurance in the case of resilience to flooding often does not correspond with other stakeholders, particularly residents (\cite{gerkensmeier_governing_2018}).

%Experience types (of flooding) - damaged belongings, survived, nuisance, direct observation, family/friend observation, heard about e.g. media

\cite{lujala_climate_2015} discusses the unexpected results that living in area with high natural hazards worsened by the changing climate did not impact peoples concern level about climate change or natural hazards in that area. However they also highlight that participants who experience direct negative experience were more concerned about the changing climate. This finding is inline with \cite{whitmarsh_are_2008}. However these findings imply that only when a place has a disaster which causes damage to individuals private property do individuals become more concerned about the changing climate or prepare themselves for increasing risk of flood damage. Another potential in \cite{lujala_climate_2015} is that participants may have been accurately ranking their personal risk, due to Norway's privileged global position. The results from this paper indicate that while their may be a decrease in social, technological and natural systems resilience to the risk of SLEs, none of the systems are completely unresilient. In general it is projected that few people will die directly from SLEs in Trondheim over the coming decades.
\paragraph{}


There is widespread insurance of buildings in Norway against damage from natural hazards (\cite{lujala_role_2020}). Households with fire insurance are required to also have insurance for the damage caused by landslides, storms and flooding. The cost of this is connected to the insurance rate for fire damage and is not dependent on the location of the house (\cite{lujala_role_2020}). The role of insurance and the welfare state in the perception of risk in Trondheim is likely significant (\cite{lujala_role_2020}), but this was not mentioned by any of the subjects regardless of their perception of the potential personal impact of flooding in these places. Future research could include questions to gain a greater understanding of how stakeholders view the role of the welfare state or insurance policies in their perception of this risk. 

\subsection{Risk Perception - Beyond SLEs}
Subjects were given the opportunity to highlight other risks to people and infrastructure which were not mentioned by the researcher. They highlighted risks associated with the natural systems and the interplay between the technological and social system. The concern about the risk of land erosion is displayed in the quote below from a subject who ranked the impact of flooding the area as significant.

\begin{itemize}
    \item "A few days ago, I saw a video of some part if Norway where an entire chunk of land just went under the water, I guess due to erosion, and I would assume that storm surges have an effect on erosion. That video was horrifying and I keep thinking, if something like that to happen here in Trondheim, which part of coast would it take away. I would love to buy/rent a house near the coast, but after watching that, I am not so sure."
\end{itemize}
\paragraph{}

At a surface level this quote highlights a fear of flooding and that they like other subjects who labelled their impact as no impact will avoid living near the coast in part due to fear of flooding. But taking it further the fear here isn't simply of SLEs but of land falling away. They are concerned about the land erosion caused by SLEs.
\paragraph{}
This fear of land falling away is brought up by several subjects at multiple points when given the opportunity to broaden the survey. For example when asked "If you would like to give other examples of risk in this area, please write here." by far the most common response was risk of landslide - often specifically due to quick clay. This response was mainly given by those taking the survey in Norwegian. 
\paragraph{}
Stakeholders awareness of SLEs appears dependant on information sources about the changing climate. Awareness appears dependant on the the stakeholders use of newspapers, family and formal education. Quick clay and landslides get media attention in Norway and this may impact awareness and resilience for these natural hazards. This could be explored in a conduction of multi-risk resilience for Trondheim.  
\paragraph{}
The other responses were about risks associated with the interplay between technological and social systems. The most common of these responses focused around lack of housing. This was sometimes discussed in the context of people having to move due to changing climate, including rising SLEs, but not always. This view is the flip side of the many respondents who stated that they will minimise personal impact of flooding due to SLEs by not residing along the coast.  As stated in the background, Trondheim has 34.6 percent of the land use in its coastal area influenced by buildings (\cite{engebakken_construction_2022}), a significant percentage which are residences. The impact to housing in Trondheim is likely not that significant within the next 70 years, but the impact in the desire to live there may be greater. This push inland could have significant economic, social and ecological impacts in Norway. The impact to the cultural systems aspect of social systems is particularly likely to change, but if this is viewed as creating a new normality, rather than as an inability to return to the previous normal this can still be considered as having high levels of resilience (\cite{cutter_place-based_2008} and \cite{cutter_community_2020}).
\paragraph{}




\section{Awareness}

\subsection{Determination of Subjects Awareness}
The results from the survey for both the variables and factors were skewed as can be seen from the figures in the section above. The results from the histogram analysis is that no variable of awareness appeared normally distributed. Upon basic transformation, including logarithmic and cubing, it was highly skewed. Furthermore, almost all factors from the results were skewed. This impacted the methods data analysis which could be used in the determination of awareness and in turn which factors impacted awareness. 

\paragraph{}
The key result from the pilot survey and focus group is that determining awareness from surveys was an appropriate method for use in determination of social resilience to SLEs in a place. Also, that utilisation of maps was not appropriate for the research sites chosen due to the low level of projected changes and that simulated images of visual extremes were a better choice.
\paragraph{}
Information access and community membership appear to be factors in the determined awareness of subjects. Over half of subjects select storm surges as a perceived risk to people and infrastructure. Subjects appear to be somewhat aware of the changing SLEs. The format in which questions of SLEs are asked appears to have an impact. 

\paragraph{}
The determination of awareness displayed in Figure 5.25 was chosen as the most appropriate variable due to the fact that only awareness questions with a simulated water level picture were included, as can be seen in the appendix. The other determination of awareness included the results from two questions which were based purely off numeric answers which gave very skewed results. When given just numeric responses, the accurate rate was non significant. For example, only 3 subjects out of 153 gave the correct answer for projections of sea level change in the last 30 years as shown in Figure 5.23.  This is unlikely given that it was multiple choice with only 7 potential answers (if split evenly then would have 21 subjects choosing this response. Figure 5.21 subjects predictions for how sea level will change in the next 30 years has more correct responses. This was only mildly significant, which combined with the lack of correct answers as shown in figure 5.22 was considered a valid reason to exclude the answers to both of these questions for creating the variable of awareness. Awareness is considered here as a continuous variable which is not independent. Rather than simply asking subjects about their awareness there was an attempt to determine the awareness. The questions used were not easy. For this reason having managed to answer some of the questions was enough for the subject to be classed as aware.

\paragraph{}
Another determination of awareness was considered which only includes results from questions on current storm surge and current high tide, excluding results of the question of the 20 year storm surge in 2090. This was considered as a potential baseline variable for current awareness to SLEs as different from awareness to future SLEs. However, the variable summary statistics are similar. For this reason Kruskal Wallis Tests were run with all factors against these two variables. Not all of these were presented as they were not all relevant for answering the hypotheses or research questions. 

\paragraph{}
The Natural Hazard to Risk to Behaviour Pipeline (Figure 4.2) shows how awareness influences behaviour. This view of awareness of natural hazard risk as being filtered through percieved distance, personal experience, values and resources dictated the search for factors which influenced awareness. 


\subsection{Visualisation of sea level change}
Discuss exclusion of questions "How much do you think the sea level will change in the next 30 years?" and "How much do you think the sea level has changed here in the past 30 years?". 

Especially why focus on answers which were picture based versus answers that were number based



\subsection{Factors Affecting Awareness}
As can be seen in the hypothesis testing summary very few factors dependent upon awareness were found. That interest level did not appear to impact awareness contrasts which contrasted with the orginal hypothesis.


The factors which were discovered that awareness is dependent upon does not mean that it is necessarily a positive relationship. Awareness is dependent upon the information source of climate of newspapers does not necessarily mean that increasing newspaper readership increases awareness of SLEs. In fact it could be the opposite. Future research can explore the relationship between these factors and awareness, perhaps by utilising linear mixed effect models.

Awareness is considered a key facet of local knowledge within the framework outlined in Figure 4.3. Other facets impacting community resilience and then social system resilience are institutions, social networks and resources. The institutions of national government, local municipalities, organisations, schools and media publications were considered within this analysis of Trondheim's resilience to SLEs. Future research could focus on better including social networks outwith institutions and the spread of the places resources. For disaster risk management in Norway responsibility lies with the municipalities and the ministry of justice and security; these institutions have high levels of resources. Trondheim by global standards has high levels of resources this will have effected the subjects view of risk and resilience as discussed on the sections about risk perception. 
\paragraph{}
The determination of awareness and the search for factors of awareness was done in such a way to attempt to start the building of a framework for quickly and cheaply determining awareness. Edited photographs, posters, social media posts and online surveys are less costly than conducting interviews, stakeholder workshops or utilising VR. Further more \cite{gerkensmeier_governing_2018} had the same results for qualitative survey as with stakeholder workshops in their investigation upon storm surges. Future research may wish to trial both methods in Trondheim to see if this is also the case here. That surveys can be used to access parts of the population who would not take part in stakeholder workshops is also a reason to include them in the understanding of resilience. However if attempting to analyse Trondheim's overall resilience rather than single-risk resilience stakeholder workshops may be more essential.


\section{Research Question 1 - How resilient will Trondheim to be to SLEs during 2022 to 2050 and 2050 to 2100? }
From the literature review Trondheim's natural systems of resilience and technological systems of resilience can be considered very high for the period of 2022 to 2050. From the results of the survey the  social systems of resilience for this period are only considered as medium as the awareness determined for the current SLEs are only somewhat. Overall Trondheim's resilience to SLEs can be considered high, due to the limited damage potential and the high level of resources available to deal with it. However, this does not mean that there is not potential economic, cultural or social impacts. Only that it is projected that the daily activities occurring in Trondheim will quickly return to normality after the likely SLE events during this period.
\paragraph{}
The awareness determined for SLEs in 2090 is higher than the awareness for 2022. This allows for the projected social system resilience for the period of 2050 to 2100 to be high. The technological system resilience is also considered as high for this period and the natural system resilience is considered high. Overall the projected resilience for Trondheim to SLEs during 2050 to 2100 is high. However, projected resilience is highly dynamic, meaning that the resilience of a place will change over time. There is requirements from the Sendai framework for and the United Nations sustainable developments goals (UN SDGs) 11 and 13 that cities resilience to natural hazards should increase (\cite{gonzalez-riancho_storm_2017}). To see if this is occurring repeated measurements of resilience is required. 

\paragraph{}
This requirement of increasing resilience brings up the question of what level of resilience should a place have. What level is acceptable to stakeholders is suggested as a method of answering this question by \cite{gerkensmeier_governing_2018}. The four research sites of Skansen, Grillstad, Brattora and Nidelva represent the most vulnerable places in Trondheim to SLEs. Taking the extreme potentials of SLEs, helps minimise the chance of overestimating the resilience Trondheim has to SLEs. The results from the survey indicate that the place had no impact on how aware subjects were to SLEs, even though each of these places had different level of natural and technological system resilience. 
\paragraph{}
This determination of Trondheims resilience to SLEs is backed up by \cite{opach_seeking_2020}. Who conducted a cluster analysis of community resilience to natural hazards for Norway and ranked Trondheim as having high economic, housing and infrastructure resilience. They were designated with lower institutional and community capital decreasing the social system resilience. However it can be argued that the way these valuations for resilience were determined were biased against cities and favoured less populous areas.



\section{Research Question 2 - Are Stakeholders aware about changes to SLEs?}

Stakeholders are somewhat aware about changes to SLEs. Whether this is acceptable level of awareness is dependent on how resilient Trondheim should be. If Trondheim is to improve its resilience to SLEs improving social system resilience should be considered as a method, as technological and natural systems are already high. The results displayed in Figures 5.24 show that when shown the information about SLEs using edited photographs and numeric valuations, 13 percent of subjects got all answers correct. 40 percent of subjects got one answer correct. Almost 40 percent of subjects got two out of the three answers correct. Only 7 percent of subject got none of the answers correct. The questions to determine awareness were deliberately difficult as there needs to be space for improvement and as a spread of the results of awareness was desired. If all subjects were classed as aware it would be impossible to determine which factors awareness is dependent upon. 
\paragraph{}
Stakeholders in Trondheim do not appear aware of the changing SLE over the past 30 years. Very few responses highlighted that sea levels in Trondheim have decreased over the last decades. This rises questions about memory. The level of difference that is required for individuals to notice and remember is particularly important here. There has been under 30 cm change in sea levels in Trondheim in the last 30 years. It is proposed that this variation was simply too small for people to notice, particularly with the continuous impacts of weather and tide. 

\paragraph{}
Stakeholders awareness could be increased, hence improving social system resilience. Future research could consider the impact of repeating this type of research on awareness of SLEs for a place. 
\paragraph{}

\section{Research Question 3 - What factors impact stakeholders' awareness of SLEs? }
The choice which format method to use while asking about SLEs is likely to have an influence on how participants perceive this change. For example considering change as metres in height vs area of land influenced may results in different perceptions of the changing risks and potential impacts.  By sticking firmly in the realm of historical fact and scientific models with these questions there has been an attempt to investigate awareness of SLEs rather than perception. How awareness was determined impacts which factors were found which awareness was dependent upon. That a small percentage were classed as fully aware will have impacts on the result of which factors impact stakeholder's awareness of SLEs. As outlined in the results chapter there were differences discovered in the results from the awareness utilising all questions and the awareness determined from questions using edited photographs. 
\paragraph{}


\subsection{Hypothesis Testing Summary}
These are the results for Research Question 3 - What factors impact stakeholders' awareness of SLEs. Two determinations of awareness were used. "Awareness All", which included all five questions from the survey upon awareness and "Awareness Photo", which only included the three questions which utilised the edited photographs showing simulated SLEs. The impact of numeric versus visual information upon SLEs is discussed in Chapter 6.
\begin{enumerate}
    \item Community membership does appear to impact stakeholders awareness of SLEs. Awareness appears dependent on the factor of residency, meaning we accept the hypothesis that residents are aware. The rest of the hypotheses dependent on community membership are rejected, apart from marine workers, which has no result as there were no responses to analyse.
  
    \item Stakeholders awareness of SLEs did not appear dependent on the factors of local knowledge considered. Awareness does not appear dependent on interest level, primary knowledge of places on reclaimed land, length of knowledge, the number of information sources about place. Whether subjects responded in Norwegian or English appears to have no impact on awareness. Stakeholders awareness does not appear dependent on their information sources about place. 

    \item  Stakeholders information sources about the changing climate does appear to impact stakeholders awareness of SLEs. Stakeholders awareness appears dependent upon their use of the information sources of family, peer reviewed publications and formal education. The awareness does not seem to be dependant on the other sources of information about the changing climate or the number of sources used. Stakeholders awareness does appear dependent on their concern about climate change. Stakeholders awareness is not dependent upon their prediction that they will be impacted by flooding from SLEs.
\end{enumerate}
\paragraph{}
How few factors were deemed to have impact on awareness was surprising. This may be due to the lack of awareness found.  Future research could flip this research question and hypotheses round and attempt to discover which factors make stakeholder's unaware of SLEs. 








